---
title: "MA678 Homework 6"
author: "Yuchen Huang"
date: "11/8/2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,out.width="0.9\\linewidth",dev="pdf",fig.align  = 'center')
library(ggplot2)
library(tidyverse)
library(knitr)
library(arm)
library(data.table)
library(foreign)
library(car)
library(faraway)
library(nnet)
library(reshape2)
library(VGAM)
library(haven)
```

## Multinomial logit

Using the individual-level survey data from the 2000 National Election
Study (data in folder NES), predict party identification (which is on a
five-point scale) using ideology and demographics with an ordered
multinomial logit model.

```{r, echo=FALSE}
NES <- read_dta("nes5200_processed_voters_realideo.dta")
## using ordered multinomial logit model
fit_polr<-polr(ordered(partyid7)~ideo+female+white+income, data=NES)
```

1.  Summarize the parameter estimates numerically and also graphically.

```{r, warning=FALSE}
display(fit_polr)

predx <- expand.grid(income = unique(NES$income), 
                    white = 1,female=0,ideo = unique(NES$ideo))
predy <- predict(fit_polr, newdata = predx,type="prob")
resd <- data.frame(predx[, c("income", "ideo","white")], party = predy) 
ggplot(melt(resd,id.var = c("income", "ideo","white")))+
    geom_bar(position = "fill", stat = "identity") + 
    aes(x=income, y = value, fill = variable) +
    facet_grid(white~ideo) +
    geom_hline(yintercept=0.5,lty=2)
```

2.  Explain the results from the fitted model.

The coefficients are ideo, female, white, income, and all of them are
statistically significant There are 6 Thresholds, the model estimates
boundaries between the ordered categories of the dependent variable
partyid7. For example, the threshold 1\|2 at 0.98 is the estimated point
in the latent variable where a respondent is equally likely to be in
category 1 or 2. The distances between these thresholds indicate how
much the latent variable needs to change to move from one category to
the next. For instance, moving from category 5\|6 to 6\|7 requires a
larger change than from 1\|2 to 2\|3. Overall, the model suggests that
ideological orientation and identifying as white are positively
associated with higher political party identification numbers, while
being female is negatively associated. Income is also positively
associated, suggesting those with higher income may identify with higher
political party identification numbers on the provided scale. The
significant coefficients and large number of observations suggest that
these findings are likely robust.

3.  Use a binned residual plot to assess the fit of the model.

C7 = party - fitted(fit_polr)[,7]-3.5)

par(mfrow = c(3,3))

```{r}
party <- NES |>
    dplyr::select(partyid7)|> na.omit()
res <- data.frame(C1 = party - fitted(fit_polr)[,1]-3.5,
                  C2 = party - fitted(fit_polr)[,2]-3.5,
                  C3 = party - fitted(fit_polr)[,3]-3.5,
                  C4 = party - fitted(fit_polr)[,4]-3.5,
                  C5 = party - fitted(fit_polr)[,5]-3.5,
                  C6 = party - fitted(fit_polr)[,6]-3.5,
                  C7 = party - fitted(fit_polr)[,7]-3.5)
par(mfrow = c(3,3))
binnedplot(fitted(fit_polr)[,1], res[,1])
binnedplot(fitted(fit_polr)[,2], res[,2])
binnedplot(fitted(fit_polr)[,3], res[,3])
binnedplot(fitted(fit_polr)[,4], res[,4])
binnedplot(fitted(fit_polr)[,5], res[,5])
binnedplot(fitted(fit_polr)[,6], res[,6])
binnedplot(fitted(fit_polr)[,7], res[,7])
```

## (Optional) Choice models

Using the individual-level survey data from the election example
described in Section 10.9 (data available in the folder NES),

```{r}

```

1.  Fit a logistic regression model for the choice of supporting
    Democrats or Republicans. Then interpret the output from this
    regression in terms of a utility/choice model.

```{r}
fit_glm <- glm(vote ~ ideo+female+white+income, data = NES)
summary(fit_glm)
```

2.  Repeat the previous exercise but now with three options: Democrat,
    no opinion, Republican. That is, fit an ordered logit model and then
    express it as a utility/choice mode

```{r}

```

## Contingency table and ordered logit model

In a prospective study of a new living attenuated recombinant vaccine
for influenza, patients were randomly allocated to two groups, one of
which was given the new vaccine and the other a saline placebo. The
responses were titre levels of hemaglutinin inhibiting antibody found in
the blood six weeks after vaccination; they were categorized as "small",
"medium" or "large".

```{=tex}
\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
treatment & small & moderate & large & Total \\ 
  \hline
placebo &  25 &   8 &   5 & 38 \\ 
 vaccine &   6 &  18 &  11 & 35 \\ 
   \hline
\end{tabular}
\end{table}
```
The cell frequencies in the rows of table are constrained to add to the
number of subjects in each treatment group (35 and 38 respectively). We
want to know if the pattern of responses is the same for each treatment
group.

```{r,echo=FALSE}
contingency <- data.frame(
    treatment = c("placebo", "vaccine"),
    small = c(25, 6),
    moderate = c(8, 18),
    large = c(5, 11),
    Total = c(38, 35))
```

1.  Using a chi-square test and an appropriate log-linear model, test
    the hypothesis that the distribution of responses is the same for
    the placebo and vaccine groups.

```{r}
chisq.test(contingency[,-1])
contingency_long <- contingency |>
    pivot_longer(cols = c(small,moderate,large,Total),
                 names_to = "category",
                 values_to = "frequency")

group1 <- glm(frequency ~ category+treatment, family = poisson(), contingency_long)
group2 <- glm(frequency ~ category*treatment, family = poisson(), contingency_long)
AIC(group1);AIC(group2)
anova(group1, group2, test = "LRT")
```

2.  For the model corresponding to the hypothesis of homogeneity of
    response distributions, calculate the fitted values, the Pearson and
    deviance residuals, and the goodness of fit statistics $X^2$ and
    $D$. Which of the cells of the table contribute most to $X^2$ and
    $D$? Explain and interpret these results.

```{r}
plot(fitted(group1), contingency_long$frequency)
resid(group1, type = "pearson")
resid(group1, type = "deviance")
```

3.  Re-analyze these data using ordered logit model (use `polr`) to
    estimate the cut-points of a latent continuous response variable and
    to estimate a location shift between the two treatment groups.
    Sketch a rough diagram to illustrate the model which forms the
    conceptual base for this analysis.

```{r}
polr(ordered(category) ~ treatment+frequency, contingency_long)
```

## High School and Beyond

The `hsb` data was collected as a subset of the High School and Beyond
study conducted by the National Education Longitudinal Studies program
of the National Center for Education Statistics. The variables are
gender; race; socioeconomic status; school type; chosen high school
program type; scores on reading, writing, math, science, and social
studies. We want to determine which factors are related to the choice of
the type of program---academic, vocational, or general---that the
students pursue in high school. The response is multinomial with three
levels.

```{r}
data(hsb)
?hsb
```

1.  Fit a trinomial response model with the other relevant variables as
    predictors (untransformed).

```{r}
fit_multi <- multinom(prog ~ gender+race+ses+schtyp+read+write+math+science+socst, data = hsb, trace = FALSE)
summary(fit_multi)
```

2.  For the student with id 99, compute the predicted probabilities of
    the three possible choices.

```{r}
id99 <- hsb[hsb$id == 99,]
predict(fit_multi, newdata = id99, type = "probs")
```

## Happiness

Data were collected from 39 students in a University of Chicago MBA
class and may be found in the dataset `happy`.

```{r}
library(faraway)
data(happy)
```

1.  Build a model for the level of happiness as a function of the other
    variables.

```{r}
fit_happy <- polr(factor(happy) ~ money+love+sex+work, data = happy)
display(fit_happy)
happy <- happy |>
    mutate(Money = scale(money, center = F),
           Work = work-3,
           Love = love-2)
fit_happyc <- polr(factor(happy) ~ Money+Love+sex+Work, data = happy)
display(fit_happyc)
```

2.  Interpret the parameters of your chosen model

By observing the coefficients, we can say that money, love, work can
improve the happiness. The p-value of sex is not statistically
significant, and the sign of sex coefficient doesn't make sense, so we
can remove this predictor.

3.  Predict the happiness distribution for subject whose parents earn
    \$30,000 a year, who is lonely, not sexually active and has no job.

```{r}
happy_pred <- predict(fit_happy, newdata = list(money = 30, sex = 0, work = 1, love = 1), type = "prob")
barplot(happy_pred)
```

## Newspaper survey on Vietnam War

A student newspaper conducted a survey of student opinions about the
Vietnam War in May 1967. Responses were classified by sex, year in the
program and one of four opinions. The survey was voluntary. The data may
be found in the dataset `uncviet`. Treat the opinion as the response and
the sex and year as predictors. Build a proportional odds model, giving
an interpretation to the estimates.

```{r}
data(uncviet)

uncviet_wider <- uncviet |>
    pivot_wider(names_from = policy,
                values_from = y)

fit_unc <- vglm(cbind(A,B,C,D) ~ sex+year, family = cumulative(parallel = T), uncviet_wider) 
summary(fit_unc)

uncviet |>
    ggplot()+
    geom_bar(position = "fill", stat = "identity")+
    aes(x = 1, y = y, fill = policy) + 
    facet_grid(sex~year)

uncviet |>
    ggplot() +
    geom_bar(position = "fill", stat = "identity")+
    aes(x = 1, y = y, fill = policy) +
    facet_grid(sex~year) +
    coord_polar("y", start = 0)

predx <- expand.grid(sex = levels(uncviet$sex), year = levels(uncviet$year))
predy <- (predict(fit_unc, newdata = predx, type = "response"))
ggplot(melt(data.frame(predx, predy), id.vars = c("sex", "year")))+
    geom_bar(stat = "identity") +
    aes(x = year, y = value, fill = variable)+
    facet_grid(~sex)
```

## Pneumonoconiosis of coal miners

The pneumo data gives the number of coal miners classified by
radiological examination into one of three categories of
pneumonoconiosis and by the number of years spent working at the coal
face divided into eight categories.

```{r}
data(pneumo, package = "faraway")
```

1.  Treating the pneumonoconiosis status as response variable as
    nominal, build a model for predicting the frequency of the three
    outcomes in terms of length of service and use it to predict the
    outcome for a miner with 25 years of service.

lines(years, predict[,1], col = "red")

lines(years, predict[,2], col = "blue")

lines(years, predict[,3], col = "green")

```{r}
counts <- xtabs(Freq ~ status+year, pneumo)
years <- pneumo$year[1:8]
fit <- multinom(t(counts) ~ years, trace = F)
summary(fit)
predict <- predict(fit, newdata = list(year = years), type = "probs")
props <- prop.table(counts,2)
plot(years, props[1,], ylim = c(0,1))
lines(years, predict[,1], col = "red")
lines(years, predict[,2], col = "blue")
lines(years, predict[,3], col = "green")
```

2.  Repeat the analysis with the pneumonoconiosis status being treated
    as ordinal.

```{r}
pneumo2 <- data.frame(status = rep(pneumo$status, pneumo$Freq), year = rep(pneumo$year, pneumo$Freq))
pneumo2$status <- ordered(pneumo2$status, levels=c("normal", "mild", "severe")) 
library(MASS)
omod <- polr(status ~ year, pneumo2)
summary(omod)

plot(years, props[1,], col="red", ylim=c(0,1))
points(years, props[2,], col="blue")
points(years, props[3,], col="green")
fitted <- predict(omod, newdata=list(year=years), type="probs") 
lines(years, fitted[,1], col="blue")
lines(years, fitted[,2], col="red") 
lines(years, fitted[,3], col="green")
```

3.  Now treat the response variable as hierarchical with top level
    indicating whether the miner has the disease and the second level
    indicating, given they have the disease, whether they have a
    moderate or severe case.

```{r}
pneumo3<- data.frame(normal=pneumo[pneumo$status == "normal","Freq"], disease=pneumo[pneumo$status == "mild","Freq"]+pneumo[pneumo$status == "severe","Freq"], mild=pneumo[pneumo$status == "mild","Freq"], severe=pneumo[pneumo$status == "severe","Freq"], year=pneumo[pneumo$status == "mild","year"])
binmodw <- glm(cbind(disease,normal) ~ year, data=pneumo3,family=binomial) 
binmodd <- glm(cbind(severe, mild) ~ year, data=pneumo3, family = binomial)

predict(binmodw,data=pneumo3,type="response")
```

4.  Compare the three analyses.

```{r}
predict(fit, newdata=list(years=25), type="probs")
predict(omod, newdata=list(year=25), type="probs")
```
